---
title: "Taxonomy formats in reference databases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taxonomy formats in reference databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Reference databases for metabarcoding encode taxonomic information in
sequence headers using different conventions. Understanding these formats
is essential when downloading databases, running taxonomic classifiers,
and summarizing results with dbpq.

## Taxonomy format overview

dbpq recognizes five taxonomy formats, grouped into two categories:
**prefix-based** formats (where each rank has a short prefix like `k__`)
and **positional** formats (where ranks are identified by their position
in a semicolon-delimited string).

### Prefix-based formats

#### UNITE format (`k__`/`p__` prefixes)

Used by: **UNITE** (general FASTA releases)

Each rank is identified by a two-letter prefix followed by double
underscores, separated by semicolons:

```
>AB123456;k__Fungi;p__Ascomycota;c__Sordariomycetes;o__Hypocreales;f__Nectriaceae;g__Fusarium;s__oxysporum
ATCGATCG...
```

| Prefix | Rank     |
|--------|----------|
| `k__`  | Kingdom  |
| `p__`  | Phylum   |
| `c__`  | Class    |
| `o__`  | Order    |
| `f__`  | Family   |
| `g__`  | Genus    |
| `s__`  | Species  |


#### SINTAX / UTAX format

Used by: **UNITE** (SINTAX release), **PR2** (UTAX format),
**MIDORI2** (SINTAX format), **Eukaryome** (SINTAX format)

Taxonomy is encoded with a `tax=` prefix, single-letter rank codes
followed by colons, separated by commas:

```
>AB123456;tax=d:Eukaryota,k:Fungi,p:Ascomycota,c:Sordariomycetes,o:Hypocreales,f:Nectriaceae,g:Fusarium,s:Fusarium_oxysporum
ATCGATCG...
```

This format is used by `vsearch --sintax` and the original USEARCH
UTAX algorithm. Note that SINTAX and UTAX use the same header format
despite being different classification algorithms.

| Prefix | Rank     |
|--------|----------|
| `d:`   | Domain   |
| `k:`   | Kingdom  |
| `p:`   | Phylum   |
| `c:`   | Class    |
| `o:`   | Order    |
| `f:`   | Family   |
| `g:`   | Genus    |
| `s:`   | Species  |


#### Greengenes2 format (`d__`/`p__` prefixes)

Used by: **Greengenes2**

Similar to the UNITE format but starts with `d__` (domain) instead
of `k__` (kingdom):

```
>abc123 d__Bacteria;p__Pseudomonadota;c__Gammaproteobacteria;o__Enterobacterales;f__Enterobacteriaceae;g__Escherichia;s__Escherichia coli
ATCGATCG...
```

| Prefix | Rank     |
|--------|----------|
| `d__`  | Domain   |
| `p__`  | Phylum   |
| `c__`  | Class    |
| `o__`  | Order    |
| `f__`  | Family   |
| `g__`  | Genus    |
| `s__`  | Species  |


### Positional formats

Some databases use semicolon-separated taxonomy without any prefix.
The meaning of each rank is determined by its position in the string.
The number of levels varies by database.

For these formats, use `rank_position` in `list_ranks_db()` to extract
a specific rank by position.

#### Unprefixed semicolon-delimited (generic)

Used by: **SILVA** (dada2-formatted), **RDP** (dada2-formatted)

```
>Bacteria;Proteobacteria;Gammaproteobacteria;Enterobacterales;Enterobacteriaceae;Escherichia;
ATCGATCG...
```

The number of levels depends on the specific training set. The
`dada2::assignTaxonomy()` classifier accepts any number of
semicolon-separated levels — with or without prefixes — via the
`taxLevels` argument.


#### PR2 format (no prefixes, 9 levels)

Used by: **PR2** (dada2 format)

PR2 uses 9 taxonomic levels specific to protist taxonomy:

```
>EU293891.1.1750_U Eukaryota;Archaeplastida;Chlorophyta;Chlorophyta_X;Mamiellophyceae;Mamiellales;Bathycoccaceae;Ostreococcus;Ostreococcus_tauri
ATCGATCG...
```

| Position | Rank         |
|----------|-------------|
| 1        | Domain       |
| 2        | Supergroup   |
| 3        | Division     |
| 4        | Subdivision  |
| 5        | Class        |
| 6        | Order        |
| 7        | Family       |
| 8        | Genus        |
| 9        | Species      |


## Sequence name placement

Beyond taxonomy encoding, formats differ in where the sequence
name (accession number, e.g. `AB123456`) appears in the FASTA header:

| Format | Sequence name location | Separator | Example header |
|--------|----------------------|-----------|----------------|
| UNITE (`k__`) | Between `>` and first `;` | `;` | `>AB123456;k__Fungi;p__Ascomycota;...` |
| SINTAX / UTAX | Between `>` and `;tax=` | `;tax=` | `>AB123456;tax=d:Eukaryota,k:Fungi,...` |
| Greengenes2 | Between `>` and first space | space | `>abc123 d__Bacteria;p__Pseudomonadota;...` |
| PR2 (positional) | Between `>` and first space | space | `>EU293891.1.1750_U Eukaryota;Archaeplastida;...` |
| dada2 (positional) | Last taxonomic level (often species) | `;` | `>Bacteria;Proteobacteria;...;Escherichia` |

In prefix-based formats (UNITE, SINTAX, Greengenes2), the sequence
name is clearly separated from the taxonomy string. In the dada2
positional format used by SILVA and RDP training sets, there is
typically no separate accession number — the header contains only
taxonomy, and the deepest level often serves as a sequence identifier.

This distinction matters when converting between formats: conversion
functions like `format2sintax()` and `format2dada2()` must correctly
extract or relocate the sequence name.


## Working with taxonomy formats in dbpq

### Detecting the format

```{r, eval = FALSE}
library(dbpq)

# Auto-detect format from file headers
detect_tax_format("my_database.fasta")
#> "unite", "sintax", "greengenes2", "pr2", or "unknown"
```

### Getting rank information

`tax_prefixes()` returns rank prefixes (character vector) for
prefix-based formats and rank positions (integer vector) for
positional formats:

```{r}
library(dbpq)

# Prefix-based formats
tax_prefixes("unite")
tax_prefixes("sintax")
tax_prefixes("greengenes2")

# Positional format (PR2)
tax_prefixes("pr2")
```

### Summarizing databases

```{r, eval = FALSE}
# UNITE format
summarize_db("unite_database.fasta", tax_format = "unite")

# SINTAX format (UNITE SINTAX, MIDORI2 SINTAX)
summarize_db("midori2_sintax.fasta", tax_format = "sintax")

# Greengenes2 format
summarize_db("greengenes2.fasta", tax_format = "greengenes2")

# PR2 positional format
summarize_db("pr2_database.fasta", tax_format = "pr2")

# Auto-detect format
summarize_db("some_database.fasta", tax_format = "auto")
```

### Listing ranks

```{r, eval = FALSE}
# Prefix-based: list phyla in a UNITE-format database
list_ranks_db("database.fasta", rank_prefix = "p__")

# Prefix-based: list phyla in a SINTAX-format database
list_ranks_db("database.fasta", rank_prefix = "p:")

# Using tax_format for convenience (extracts first rank)
list_ranks_db("database.fasta", tax_format = "unite")

# Positional: list genera in a PR2 database (position 8)
list_ranks_db("database.fasta", tax_format = "pr2", rank_position = 8)

# Positional: list phyla in any unprefixed database (position 2)
list_ranks_db("database.fasta", rank_position = 2)
```

### Converting between formats

```{r}
# UNITE (k__) → SINTAX
format_fasta_db(
  taxnames = "AB123;k__Fungi;p__Ascomycota;c__Sordariomycetes",
  output_format = "sintax"
)

# SINTAX → UNITE (k__)
format_fasta_db(
  taxnames = "AB123;tax=k:Fungi,p:Ascomycota,c:Sordariomycetes",
  output_format = "unite"
)

# Greengenes2 → dada2
format_fasta_db(
  taxnames = "abc123 d__Bacteria;p__Pseudomonadota;g__Escherichia",
  output_format = "dada2"
)
```

## Downloading databases in specific formats

Several databases offer SINTAX-formatted downloads alongside their
default format:

```{r, eval = FALSE}
# UNITE: default vs SINTAX
download_unite_db(dest_dir = "databases")
download_unite_db(dest_dir = "databases", taxonomic_format = "sintax")

# PR2: dada2 vs UTAX/SINTAX
download_pr2_db(dest_dir = "databases", format = "dada2")
download_pr2_db(dest_dir = "databases", format = "sintax")

# MIDORI2: multiple formats available
download_midori2_db(gene = "CO1", format = "dada2")
download_midori2_db(gene = "CO1", format = "SINTAX")
```

For databases that only provide one format, use `format_fasta_db()` to convert:

```{r, eval = FALSE}
# Download in dada2 format, then convert to SINTAX
download_silva_db(dest_dir = "databases", format = "dada2")
format_fasta_db(
  fasta_db = "databases/silva_nr99_v138.2_toSpecies_trainset.fa.gz",
  output_format = "sintax",
  output_path = "databases/silva_sintax.fasta"
)
```


## Database formats and classification algorithms

Different taxonomic classifiers expect different input formats.
The table below shows which formats work with which classifiers,
including those available via `MiscMetabar::add_new_taxonomy_pq()`:

| Classifier | Expected format | dbpq download | `add_new_taxonomy_pq()` method |
|------------|----------------|---------------|-------------------------------|
| `dada2::assignTaxonomy()` | Any `;`-separated taxonomy | `format = "dada2"` | `method = "dada2"` |
| `dada2::addSpecies()` | `ID Genus species` | `format = "dada2_species"` | `method = "dada2_2steps"` |
| `vsearch --sintax` | SINTAX (`tax=`) | `taxonomic_format = "sintax"` | `method = "sintax"` |
| VSEARCH LCA (`--usearch_global`) | Any FASTA | Any | `method = "lca"` |
| IDTAXA | Custom training set | Convert with `format2dada2()` | `method = "idtaxa"` |
| BLASTn | Any FASTA | Any | `method = "blastn"` |

### Choosing a database and classifier

**For ITS (fungi):**

- UNITE + dada2: `download_unite_db()` + `method = "dada2"`
- UNITE SINTAX + vsearch: `download_unite_db(taxonomic_format = "sintax")` + `method = "sintax"`

**For 16S (bacteria/archaea):**

- SILVA: `download_silva_db()` + `method = "dada2"`
- RDP: `download_rdp_db()` + `method = "dada2"`
- Greengenes2: `download_greengenes2_db()` + `method = "dada2"`

**For 18S (protists/eukaryotes):**

- PR2: `download_pr2_db()` + `method = "dada2"` or `"sintax"`
- Eukaryome: `download_eukaryome_db()`

**For COI (metazoa):**

- MIDORI2: `download_midori2_db()`
- BOLD: `download_bold_db()`

**For rbcL (diatoms):**

- Diat.barcode: `download_diatbarcode_db()` or `diatbarcode` R package

**For 18S AMF (arbuscular mycorrhizal fungi):**

- MaarjAM: `download_marjaam_db()`
